A Simple Ratio for Estimating the Effective Reproduction Value of an Epidemic

### 2021-01-16 docs/manuscript/manuscript.Rmd
update manuscript

From a public health perspective, the $R_t$ can be a method to communicate whether an epidemic is growing in a region. However, this communication may be hindered by the fact that many of the current methods for estimating $R_t$ require advanced mathematics to understand and to apply [@cori_new_2013; @wallinga_different_2004; @wallinga_how_2007]. Typically these methods require a description of the generation interval distribution, which is the distribution of times between when a primary case becomes infectious to when a secondary case becomes infectious.

@@

```{=tex}
\begin{equation}
  R_t \approx \frac{I_t}{I_{t-d}}
  (\#eq:simple-ratio-eq)
\end{equation}
```
## A Simple Ratio for Estimating Rt {.unnumbered}

As described by previous publications [@wallinga_how_2007], we can define a function $n(a)$ that describes the expected number of new cases produced today (time $t$) by a single person who was infected $a$ days ago (at time $t-a$). Therefore, to calculate $I_{ta}$ (the expected number of new infections created today by all individuals who became infected $a$ days ago), we multiply $n(a)$ by the number of new infections created $a$ days ago ($I_{t-a}$).

```{=tex}
\begin{equation}
  I_{ta} = I_{t-a}n(a)
  (\#eq:eq2)
\end{equation}
```
Then to determine the total expected number of new infections today ($I_t$), we sum $I_{ta}$ for all values of $a$, as shown in \cref{eq:renewal-eq-1,eq:renewal-eq-2}. In other words, $I_t$ can be calculated by taking all individuals infected prior to $t$, calculating the expected number of new infections each one will produce based on the age of their infection, and summing all those values together.

```{=tex}
\begin{equation} 
  I_t = \sum_{a=0}^\infty{I_{ta}}da
  (\#eq:renewal-eq-1)
\end{equation}
```
```{=tex}
\begin{equation} 
  I_t = \sum_{a=0}^\infty{I_{t-a}n(a)}da
  (\#eq:renewal-eq-2)
\end{equation}
```
Next we can substitute the generation interval function ($g(a)$) into \cref{eq:renewal-eq-2} by recognizing that the generation interval distribution is equal to $n(a)$ normalized by the total number of infections a single infected person is expected cause (i.e. $R_t$).

```{=tex}
\begin{equation} 
  g(a) = \frac{n(a)}{R_t}
  (\#eq:generation-interval-eq)
\end{equation}
```
```{=tex}
\begin{equation} 
  I_t = \sum_{a=0}^\infty{I_{t-a}g(a)R_t}da
  (\#eq:sub-in-rt-1)
\end{equation}
```
```{=tex}
\begin{equation} 
  \frac{I_t}{R_t} = \sum_{a=0}^\infty{I_{t-a}g(a)}da
  (\#eq:sub-in-rt-2)
\end{equation}
```
```{=tex}
\begin{equation} 
  R_t = \frac{I_t}{\sum{I_{t-a}g(a)}da}
  (\#eq:sub-in-rt-3)
\end{equation}
```
If we make the assumption that there is minimal variance in the generation interval distribution (i.e. $g(a)$ equals approximately 1 when $a$ equals the mean generation interval ($d$) and $g(a)$ equals approximately zero for all other values of $a$), then Equation \cref{eq:sub-in-rt-3} can be simplified to:

@@

```{=tex}
\begin{equation} 
  R_t \approx \frac{\sum_{t - \frac{\tau}{2}}^{t + \frac{\tau}{2}}I_i}{\sum_{t - \frac{\tau}{2}}^{t + \frac{\tau}{2}}I_{i-d}} = \frac{n_{t}}{n_{t-d}}
  (\#eq:eq6)
\end{equation}
```
A confidence interval can be generated for this ratio by carrying forward a confidence interval in the generation interval.

@@

## Rt Estimation {.unnumbered}

@@

Over the `r simulation_parameters$Nsims` simulation, the median estimation error of the daily $R_t$ by the simple ratio method was `r error_median` (90% range of `r error_quantile_025` to `r error_quantile_975`) (\cref{fig:sim-results}**B**). The simple ratio has a tendency to over-estimate the $R_t$, especially as the true value of $R_t$ increases above 1.5 (\cref{fig:sim-results}**C**). This is consistent with previous mathematical analysis of this simplified method of estimating $R_t$ [@wallinga_how_2007].

(ref:sim-results-cap) \textbf{The Simple Ratio approximates the effective reproduction in simualted epidemics.} One hundred epidemics were simulated using an SEIR model and the simple ratio and Cori method were used to estimate the known $R_t$ over time for each simulated epidemic. \textbf{(A)} The true $R_t$ (black line) and estimated $R_t$ (colored lines) over time for one of simulated epidemics. \textbf{(B)} The distribution of the absolute errors of both estimation methods across all simulated epidemics. \textbf{(C)} The estimation error of the simple ratio method is associated with size of true $R_t$

```{r sim-results, fig.cap = '(ref:sim-results-cap)'}
# Plots
panel_A <- rt_estimates %>%
  filter(simulation_run == 66) %>%
  create_estimate_plot()

@@

```{r estimation-inaccParams,  fig.cap = "(A) estimation with gi mean mis-specified. (B) estimation with gi sd mis-specified."}

>>>>

From a public health perspective, the $R_t$ can be a method to communicate whether an epidemic is growing in a region. However, this communication may be hindered by the fact that many of the current methods for estimating $R_t$ require advanced mathematics to understand and to apply [@cori_new_2013; @wallinga_different_2004; @wallinga_how_2007]. Typically these methods require a description of the generation interval distribution, which is the distribution of times between when a primary case becomes infected to when a secondary case becomes infected.

@@

```{=tex}
\begin{equation}
  R_t \approx \frac{I_t}{I_{t-d}}
  (\#eq:simple-ratio-eq)
\end{equation}
```
## Deriving the Simple Ratio for estimating $R_t$ {.unnumbered}

A subset of previous methods for estimating $R_t$ are based on the following relationship:

```{=tex}
\begin{equation} 
  R_t = \frac{I_t}{\sum_{a=0}^\infty{I_{t-a}g(a)}}
  (\#eq:original-rt)
\end{equation}
```
$I_t$ is the number of new cases produced on day $t$; $g(a)$ is the infectiousness of a person infected $a$ days ago - $g(a)$ can be thought of as equivalent to to the generation interval distribution. To understand the derivation of this equation, we refer readers to [@wallinga_how_2007].

If we make the simplifying assumption that there is minimal variance in $g(a)$ (i.e. $g(a)$ is 1 when $a$ equals the mean generation interval ($d$) and $g(a)$ is zero for all other values of $a$), then Equation \cref{eq:original-rt} can be simplified to:

@@

```{=tex}
\begin{equation} 
  R_t \approx \frac{\sum_{t - \frac{\tau}{2}}^{t + \frac{\tau}{2}}I_i}{\sum_{t - \frac{\tau}{2}}^{t + \frac{\tau}{2}}I_{i-d}} = \frac{n_{t}}{n_{t-d}}
  (\#eq:eq6)
\end{equation}
```
A confidence interval can be generated for this ratio by carrying forward a confidence interval of the mean generation interval.

@@

## $R_t$ Estimation {.unnumbered}

@@

Over the `r simulation_parameters$Nsims` simulation, the median estimation error of the daily $R_t$ by the simple ratio method was `r error_median` (95% range of `r error_quantile_025` to `r error_quantile_975`) (\cref{fig:sim-results}**B**). The simple ratio has a tendency to over-estimate the $R_t$, especially as the true value of $R_t$ increases above 1.5 (\cref{fig:sim-results}**C**). This is consistent with previous mathematical analysis of this simplified method of estimating $R_t$ [@wallinga_how_2007]. When $R_t$ is relatively stable, the simple method may over-estimate $R_t$; however during periods of rapid change in $R_t$ (e.g. going in and out of lockdown), the simple method may either over- or under-estimate $R_t$ (\cref{fig:sim-results}**D**).

(ref:sim-results-cap): \textbf{The Simple Ratio approximates the effective reproduction in simualted epidemics.} One hundred epidemics were simulated using an SEIR model and the simple ratio and Cori method were used to estimate the known \$R_t\$ over time for each simulated epidemic. \textbf{(A)} The true \$R_t\$ (black line) and estimated \$R_t\$ (colored lines) over time for one of simulated epidemics. \textbf{(B)} The distribution of the absolute errors of both estimation methods across all simulated epidemics. \textbf{(C)} The estimation error of the simple ratio method is associated with size of true \$R_t\$. \textbf{(D)} The estimation error of both methods over time for the simulated epidemics. Each line represents the error in \$R_t\$ estimation for a given simulation. Note that for both methods, a substantial proportion of error occurs around days 60 and 130 when there was is rapid change in \$R_t\$

```{r sim-results, fig.cap = '(ref: sim-results-cap)'}

# Plots
panel_A <- rt_estimates %>%
  filter(simulation_run == 66) %>%
  create_estimate_plot()

@@

```{r estimation-inaccParams,  fig.cap = "\\textbf{$R_t$ estimation error when the mean or standard deviation of the generation interval is mis-specified}. Similar to \\textbf{Figure 1}, this figure demonstrate the distribution of absolute error when estimating $R_t$ for 100 simulated epidemics.  \\textbf{(A)} Distribution of errors when the mean of the generation interval is mis-specified as 2 days too long. The average \\textbf{(B)} Distribution of errors when the standard deviation is mis-specified as 3 days too long."}
